---
title: "mod4_hw1_st525_oursop"
author: "Philip Ourso"
date: "10/23/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(tidytext)
library(directlabels)
library(colorspace)
library(hexbin)
library(RColorBrewer)
library(vtable)
library(km.ci)
```
## Module 4 HW1

    1. In the definition of...
  
      
      a. What does the term di/ ni estimate?
          The term estimates the conditional probability of failure at time ti given that the individual was alive just before time ti.
      
      b. What does the term 1 - di/ ni estimate?
          The term estimates the conditional probability of survival of time ti given that the individual was alive just before time ti.
          
      c. Do the exact values of censoring times play any role in the Kaplan-Meier estimator?
          The exact time of censoring can't produce a change in the KM estimator, but it does affect the number of survivors upon the subsequent event time. 
          
      d. Do the exact values of event times play any role in the Kaplan-Meier estimator?
          Event times indicate the only points at which the KM estimator decreases.
          
      e. Show that when there is no censoring, the Kaplan-Meier estimator S_hat(t) reduces to the empirical survival function.
          Several rows of survival data are displayed below, with both the Kaplan-Meier estimator and the empirical survival function calculated, assuming no censoring. As can be seen, the two estimators are in agreement.
      
          ti| di| ni | 1-di/ni | KM                | S_hat(ti)     |
          -----------------------------------------------------------
          6 | 3 | 21 | 0.857   |0.857              | 18/21 == 0.857|
          7 | 1 | 18 | 0.944   |0.857*0.944==0.809 | 17/21 == 0.809|
          10| 1 | 17 | 0.941   |0.809*0.941==0.761 | 16/21 == 0.762|
      
    2. Suppose that the following remission...
        3.0, 6.5, 6.5, 10, 12, 15, 8.4+, 4.0+, 5.7+, 10+
      a. Compute the Kaplan-Meier estimate manually and plot the estimated survival curve. (In your solution, please include any tables that you used to construct the Kaplan-Meier).
          ti  | di| ni | 1-di/ni | KM                |
          --------------------------------------------
          3   | 1 | 10 | 0.9     |0.9                |
          4+  |   |  9 |         |                   |
          5.7+|   |  8 |         |                   |
          6.5 | 2 |  7 | 0.714   |0.9*0.714 == 0.643 |
          8.4+|   |  5 |         |                   |
          10  | 1 |  4 | 0.750   |0.643*0.750== 0.482|
          10+ |   |  3 |         |                   |
          12  | 1 |  2 | 0.500   |0.482*0.5 == 0.241 |
          15  | 1 |  1 | 0       |0.241*0 == 0       |

```{r, echo=FALSE, warning=FALSE, message=FALSE}
df = data.frame(
  ti = c(0, 3, 6.5, 10, 12, 15),
  km = c(1, 0.9, 0.643, 0.482, 0.241, 0)
)
#head(df)

ggplot(df, aes(x=ti, y=km)) +
  geom_step() +
  geom_hline(yintercept=0.5, color='red', linetype='dashed') +
  theme_minimal() +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    title = "Kaplan-Meier Estimator",
    x = "Event time (months)",
    y = "Estimate survival function"
  )

```
     
      b. Find the median survival time and the corresponding estimated survival function value.
          The median survival time is defined as the smallest time at which the survival probability drops to 0.5 or below. It can be seen that the median survival time is 10 months, with an estimated survival function value of 0.482.
        
      c. Compute the cumulative hazard at 10 months.
          H(t) = -ln( S(t)) = -ln( 0.482) = 0.73
        
    3. The following data is based on a...
      a. Compute the nonparametric estimate for the survival function manually using Life-Table method. 
      
          ei = ni - ci/2
          mi = di/ ei
          Si = Si-1 * 1 - mi

          Age int  | di | ci | ei     | mi   | Si
          -------------------------------------------------
          45-50    | 17 | 29 | 1556.5 |0.011 | 0.99       |
          50-55    | 36 | 60 | 1509.5 |0.024 | 0.97       |
          55-60    | 62 | 83 | 1432.0 |0.043 | 0.92       |
          60-65    | 76 |441 | 1149.5 |0.066 | 0.86       |
          65-70    | 50 |439 |  854.0 |0.059 | 0.81       |
          70-75    |  9 |262 |  673.0 |0.013 | 0.80       |
          75-80    |  0 |  7 |  660.5 |    0 | 0.80       |

      b. What is the estimated survival probability for interval [55-60]?
          0.92
          
      c. Provide an interpretation of the survival probability in (b).
          There is a 0.92 probability that an individual will not develop CHD between ages 55 and 60 given that they've not developed CHD by age 55.

    4. Here we consider the pharmacoSmoking data set...
      a. Import the data into SAS or R and calculate the survival time of interest as the number of days between admission date and last follow up date.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df = read.csv("pharmacoSmoking-new.csv",
              header = TRUE,
              colClasses = c(rep(NA, 5), "factor", rep(NA, 4), "character", "character", NA, NA)
              )
df = df %>%
  mutate( gender = ifelse(gender==0, "Female", "Male"))

df$admittime = df$admitdate %>%
  as.Date(format = c("%m/%d/%Y"))

df$ftime = df$fdate %>%
  as.Date(format = c("%m/%d/%Y"))

df$time = difftime(df$ftime, df$admittime, units = 'days') %>%
  as.numeric() 

head(df)
```

      b. Compute the Kaplan-Meier estimator for male and female groups separately and plot the two survival curves in the same figure.
      
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(survival)
library(survminer)

censored <- Surv(df$time, df$relapse) #Creates a survival object

fw <- survfit(Surv(time, relapse) ~ 1, data = df[df$gender=="Female",], conf.type="none") 

fm <- survfit(Surv(time, relapse) ~ 1, data = df[df$gender!="Female",], conf.type="none") 


par(cex=.7)
plot(fw,xlab="Time",ylab="Relapse probability", col=4)
lines(fm, lty=2,col=2)
legend(0.5, .2, c("Female", "Male" ), lty=c(1,2),col=c(4,2))

#fit = list(male=fm, female=fw)

#table1 <- summary(f1, times = c(0, unique(MP6$time6mp)))
#censored <- Surv(MP6$time6mp, MP6$cstatus) #Creates a survival object
#ggsurvplot(fit = fit, data = censored, combine = TRUE, 
#conf.int = FALSE, conf.int.style = 'step',
#legend.title = 'Kaplan-Meier plot with point-wise confidence intervals')
```
      
      c. Can an Exponential distribution fit the survival time well for male or female group?
          The -log(survival time) vs time plots are approximately linear, so its reasonable to model the survival time as an exponential distribution.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
table_w <- summary(fw, times = c(0, unique(df[df$gender=="Female","time"])))

# times is the vector on which you want to get the estimated survival function values.
table_w <- data.frame(
  time = table_w$time, 
  survival = table_w$surv, 
  failure = 1-table_w$surv, 
  Survival.Std.Err = table_w$std.err, 
  No.Left = table_w$n.risk, 
  No.Failed = table_w$n.event, 
  No.Censored = table_w$n.censor
  )

# Extract all information we need to build a data frame.
print(table_w[, 1:7], row.names = FALSE)

table_w2 <- table_w[c(table_w$No.Failed > 0), ]
table_w2 <- data.frame(rbind(table_w[1, ], table_w2, table_w[17, ]))
# I do the data cleaning here. Extract all event happening rows out of table 1 and insert 
# the first and the last row to build the table2.

ggplot(data = table_w2) + 
  geom_line(aes(x = time, y = -log(survival))) +
  geom_point(aes(x = time, y = -log(survival)), size = 2, shape = 1) +
  labs(title = 'Negative Log of Estimated Female Survivor Function') +
  ylab('-log(Survival Probability)') + xlab('time') +
  scale_x_continuous(limits = c(0, 35))

table_m <- summary(fw, times = c(0, unique(df[df$gender!="Female","time"])))

# times is the vector on which you want to get the estimated survival function values.
table_m <- data.frame(
  time = table_m$time, 
  survival = table_m$surv, 
  failure = 1-table_m$surv, 
  Survival.Std.Err = table_m$std.err, 
  No.Left = table_m$n.risk, 
  No.Failed = table_m$n.event, 
  No.Censored = table_m$n.censor
  )

# Extract all information we need to build a data frame.
print(table_m[, 1:7], row.names = FALSE)

table_m2 <- table_m[c(table_m$No.Failed > 0), ]
table_m2 <- data.frame(rbind(table_m[1, ], table_m2, table_m[17, ]))
# I do the data cleaning here. Extract all event happening rows out of table 1 and insert 
# the first and the last row to build the table2.

ggplot(data = table_m2) + 
  geom_line(aes(x = time, y = -log(survival))) +
  geom_point(aes(x = time, y = -log(survival)), size = 2, shape = 1) +
  labs(title = 'Negative Log of Estimated Male Survivor Function') +
  ylab('-log(Survival Probability)') + xlab('time') +
  scale_x_continuous(limits = c(0, 35))
```
        
        d. For both male and female groups, plot the 95% confidence interval for the survival curves using linear transformation method and log-transformed method
        
```{r}
censored <- Surv(df$time, df$relapse) #Creates a survival object

# female survival curve and CIs
fw <- survfit(Surv(time, relapse) ~ 1, data = df[df$gender=="Female",], conf.type="none") 
#fw.ci.log<-survfit(Surv(time, relapse) ~ 1, data = df[df$gender=="Female",], conf.type = 'log')
fw.ci.log = km.ci(fw, conf.level = 0.95, method = "log")
#fw.ci.linear<-survfit(Surv(time, relapse) ~ 1, data = df[df$gender=="Female",], conf.type = 'plain')
fw.ci.linear = km.ci(fw, conf.level = 0.95, method = "linear")

fm <- survfit(Surv(time, relapse) ~ 1, data = df[df$gender!="Female",], conf.type="none") 
#fm.ci.log<-survfit(Surv(time, relapse) ~ 1, data = df[df$gender!="Female",], conf.type = 'log')
fm.ci.log = km.ci(fm, conf.level = 0.95, method = "log")
#fm.ci.linear<-survfit(Surv(time, relapse) ~ 1, data = df[df$gender!="Female",], conf.type = 'plain')
fm.ci.linear = km.ci(fm, conf.level = 0.95, method = "linear")


par(cex=.7)
plot(fw,xlab="Time",ylab="Relapse probability", col=4)
lines(fw.ci.log, lty=1,col=4)
lines(fw.ci.linear, lty=3,col=4)
lines(fm, lty=2,col=2)
lines(fm.ci.log, lty=2,col=2)
lines(fm.ci.linear, lty=3,col=2)
legend(0.5, .2, c("Female", "Male" ), lty=c(1,2),col=c(4,2))


#ci.log<-survfit(Surv(time6mp, cstatus) ~ 1, data = MP6, conf.type = 'log')
#ci.linear<-survfit(Surv(time6mp, cstatus) ~ 1, data = MP6, conf.type = 'plain')
#par(cex=.7)
#plot(f1,xlab="Time",ylab="Survial probability",lwd=2)
#lines(ci.log, lty=2,col=2)
#lines(ci.linear, lty=4,col=3)

```
        
        e. From the estimated survival curves, how do the survival experiences of the female and male patients compare?
            Females initially show a better survival probability, and indeed have a better median survival time, but begin to exhibit a similar survival time, until about 60 days or so, after which males exhibit a better survival probability.
        
### Appendix: R code
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
