---
title: "mod7_hw1_st525_oursop"
author: "Philip Ourso"
date: "11/14/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(colorspace)
library(directlabels)
library(hexbin)
library(km.ci)
library(RColorBrewer)
library(survival)
library(survminer)
library(tidyverse)
library(tidytext)
library(vtable)

```
## Module 7 HW1
  
1. A randomized clinical trial studies survival times...  
  a. Use the Kaplan-Meier method to estimate the survival functions for the two groups separately, and plot two Kaplan-Meier curves in the same figure.  
    The KM estimator survival functions are plotted below.  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_1 = data.frame(
  time = c(c(10, 10, 12, 12, 13, 14, 17, 20, 26, 34, 35, 37, 38, 39, 40, 40), 
           c(7, 9, 9, 10, 11, 11, 13, 14, 14, 15, 16, 17, 19, 21, 22, 27)),
  status = c(rep(1,1), rep(0,1), rep(1,1), rep(0,1), rep(1,2), rep(0,1), rep(1,2), rep(0,1), rep(1,5), rep(0,1), rep(1,2), rep(0,1), rep(1,2), rep(0,1), rep(1,2), rep(0,1), rep(1,4), rep(0,1), rep(1,1), rep(0,1)),
  group = c(rep(0,16), rep(1,16)))

#head(df_1)

df_A = df_1 %>%
  filter(group == 0 )

fit1 <- survfit(Surv(time, status) ~ 1, data = df_A)
ggsurvplot(fit = fit1, data = df_A, conf.int = TRUE) +
  labs(
    title="KM Estimator Survival Function for Treatment Group"
  )

df_B = df_1 %>%
  filter(group == 1 )

fit2 <- survfit(Surv(time, status) ~ 1, data = df_B)
ggsurvplot(fit = fit2, data = df_B, conf.int = TRUE) +
  labs(
    title="KM Estimator Survival Function for Placebo Group"
  )

```  
  
  b. Now perform a log-rank test for treatment effect. Please report your test statistic and a p-value. Does the treatment have an effect on patient's survival, based on the log-rank test?  
    The log-rank test produces a test statistic of 5.2 with a corresponding p-value of 0.02, providing strong evidence of a difference in survival times between the two groups and leading us to reject the null hypothesis.  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#fit_b <- survfit(Surv(time, status) ~ group, data = df_1)
#ggsurvplot(fit = fit_b, data = df_1, conf.int = TRUE)
logrank<-survdiff(Surv(time, status) ~ group, data = df_1)
logrank
```  
  
  c. Assume that the survival times for both groups follow exponential distribution, but with possibly different parameters. Test the hypothesis of no treatment effect using Wald test.  
    The Wald test statistic is 1.5 with a p-value of 0.22, leading us to fail to reject the null hypothesis of no treatment effect. This is different from our log-rank test result.  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
fit_c <- survreg(formula = Surv(time, status) ~ 
                  group,
                data = df_1, dist = 'exponential')
summary(fit_c)
```  
  
  d. Discuss the assumptions underlying these two approaches, and the reasons for the discrepancy between two testing procedures.  
    The log-rank test is a non-parametric approach and makes no assumptions about the underlying distributions at work, instead assuming that if there were no differences between groups then events would happen proportionally over time. The test can suffer reduced power should the survival curves fail to exhibit proportional hazard rates, which doesn't appear to be the case here.  
    The exponential regression model is both an accelerated failure time model and a proportional hazards model, assuming a log-linear model of survival time with survival times following an exponential distribution and hence the model's error term following a standard extreme value distribution. A constant hazard is when applying such models.  
    If the survival times in fact follow a different distribution, or the log survival times are non-linear, then the exponential regression model would not be a good fit. 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
fit_b <- survfit(Surv(time, status) ~ group, data = df_1)
ggsurvplot(fit = fit_b, data = df_1, conf.int = TRUE)
```
  
2. In a study to assess the time to first exit-site infection...  
  a. Fit a Weibull model to patients with percutaneous placement.  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_2  = read.csv("kidney.txt", sep = " ")
df_2$group = factor(ifelse(df_2$Type==2, 0, 1))
#head(df_2)

fit_2a = survreg(formula = Surv(Time, Delta) ~ 
                  1,
                data = df_2[df_2$Type==2,], dist = 'weibull')
summary(fit_2a)
exp(0.616)*0.265
```  
  
  b. Find the maximum likelihood estimates of scale parameter and its standard error.  
    From above, the scale parameter is 1.85 and its standard error is calculated as 0.4906.  
  
  c. Test the hypothesis that the shape parameter is equal to 1 using both the likelihood ratio test and the Wald test. Report your test statistics and p-values.  
    Using the Likelihood ratio test to compare Weibull and exponential regression models produces a test statistic of 7.13 with a corresponding p-value of 0.008, leading us to reject the null hypothesis of a simpler model, so we retain the more complex Weibull regression model.  
    Using the Wald test to test if scale == 1, or equivalently if log(scale) == 0, produces a test statistic of 5.4 with a corresponding p-value of 0.02, again leading us to reject the null hypothesis.  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# likelihood ratio test -> fit Weibull and exponential
fit_2c = survreg(formula = Surv(Time, Delta) ~ 
                  1,
                data = df_2[df_2$Type==2,], dist = 'exponential')
2*(fit_2a$loglik[1] - fit_2c$loglik[1])
1 - pchisq(2*(fit_2a$loglik[1] - fit_2c$loglik[1]),1)

# wald test for log(scale)==0
(0.616/0.265)^2
1-pchisq((0.616/0.265)^2,1)


```  
  
  d. What is your conclusion based on your test results in (c)?  
    For those patients with percutaneous placement, the exponential model is insufficient and we should retain the Weibull regression model.  
  
  e. Fit a Weibull regression model to this data with a single covariate, Z, that indicates group membership.  
  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
fit_2e = survreg(formula = Surv(Time, Delta) ~ group,
                data = df_2, dist = 'weibull')
summary(fit_2e)
```  
  
  f. Find the maximum likelihood estimates of the regression coefficient of Z and its standard error. Provide an interpretation of the regression coefficient.  
    Redefining the group parameter as a factor in which group == 0 signifies percutaneous placement and group == 1 signifies surgical placement results in a group == 1 coefficient of -0.623 with a standard error of 0.469. The negative coefficient indicates that group 1 membership (surgical placement) has a detrimental impact on survival time; in other words group 1 members can expect ~46% decrease in time to first exit-site infection with respect to group 0 members. 
    It should be noted that the corresponding p-value is 0.18 and hence the coefficient is not found to be statistically significant.  
  
  g. Test the hypothesis of no effect of catheter placement on the time to infection using both likelihood ratio test and Wald test. Report you test statistics and p-values.  
    The likelihood ratio test statistic is 1.8 with a corresponding p-value of 0.18.  
    The Wald test statistic is 1.76 with a corresponding p-value of 0.18.  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(fit_2e)
fit_2g = survreg(formula = Surv(Time, Delta) ~ 1,
                data = df_2, dist = 'weibull')
summary(fit_2g)

# likelihood ratio test
lkl_2g = 2*(-122-(-122.9))
lkl_2g
1 - pchisq(lkl_2g,1)

# Wald
wld_2g = (-0.623/0.469)^2
wld_2g
1 - pchisq(wld_2g, 1)
```  
  
  h. Does the Weibull regression model provide a good fit to the data? Justify your conclusion using an appropriate residual plot.  
    Reviewing the negative log plot of the Survival function of Cox-Snell residuals against the residuals themselves exhibits an approximately linear plot with slope 1 passing through the origin, indicating the model is a good fit to the data.  
```{r}
csr = df_2$Time * exp(-fit_2g$linear.predictors) 
#qplot(df_2$Time, csr, main = c('Cox-Snell residuals from Weibull regression')) +
#  xlab('time') + ylab('Cox-Snell Residual') 

fit_2h = survfit(Surv(csr, Delta) ~ 1, data=df_2)
plot(fit_2h$time, -log(fit_2h$surv))
```

3. Here we consider the pharmacoSmoking dataset...
  a. Report the estimated coefficients with p-values. Do any of these covariates have significant effects on survival time?
    None of the covariates appear to have a statistically significant effect on survival time.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
df_3 = read.csv("pharmacoSmoking-new.csv",
              header = TRUE,
              colClasses = c(rep(NA, 5), "factor", rep(NA, 4), "character", "character", NA, NA)
              )
df_3 = df_3 %>%
  mutate( grp = factor(grp), 
          employment = factor(employment))

df_3$admittime = df_3$admitdate %>%
  as.Date(format = c("%m/%d/%Y"))

df_3$ftime = df_3$fdate %>%
  as.Date(format = c("%m/%d/%Y"))

df_3$time = difftime(df_3$ftime, df_3$admittime, units = 'days') %>%
  as.numeric() 

df_3 = df_3[df_3$time>0,]
#dim(df)

fit_3a <- survreg(formula = Surv(time, relapse) ~ age + yearsSmoking + priorAttempts + gender + employment,
                data = df_3, dist = 'weibull')
summary(fit_3a)
```  
  
  b. Interpret the regression coefficients for Age, yearsSmoking, priorAttempts, gender, employment.  
    Age has a coefficient of 0.06, indicating that for each year increase the survival time increases exp(0.06)-1~ 0.06%.  
    yearsSmoking indicates that for each additional year smoking decreases survival time ~1%.  
    priorAttempts indicates nearly no change for each additional prior attempt at quitting.  
    gender indicates males have a ~32% increase in survival time relative to females.  
    employment indicates that part-time employees and "other" employees have a decrease of ~64% and ~54% respectively relative to full-time employees.  
  
  c. Is the Weibull regression a good fit to the data?  
    Reviewing the Cox-Snell residual plot for model diagnosis reveals an approximately linear plot through the origin. The slope is nearly 1, so its not unreasonable to model the survival times with a Weibull regression model.  
```{r}
csr_3 = df_3$time * exp(-fit_3a$linear.predictors) 

fit_3c = survfit(Surv(csr_3, relapse) ~ 1, data=df_3)
plot(fit_3c$time, -log(fit_3c$surv))

```  
  
  d. If the Weibull regression is not appropriate for the data set, discuss how it will affect your estimation and interpretation results in (a) and (b).  
    Had the Weibull regression model been a poor fit for the data it would call into question the conclusions reached in earlier questions. The effect of covariates upon survival time, as indicated by the estimated coefficients, would not be trusted. Furthermore, the statistical significance of the covariates' coefficients would also be in question.    
  
  

### Appendix: R code
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
